Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-04-29 paper/main.tex
more revisions

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{amsmath, amsfonts, amssymb, mathrsfs, dsfont}
\usepackage{dcolumn}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{filemod}
\usepackage{natbib}
\usepackage[ruled, vlined]{algorithm2e}
\usepackage{floatrow}
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{xcolor} % only for coloring text while revising

@@

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and which are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all available simulation information into an effective estimate of the true underlying
matter spectrum.

There are several challenges to estimating the underlying spectra.  Our model
must: handle multiple correlated functional observations (like the low resolution
runs shown in Figure \ref{fig:plot_data}), offer enough 
flexibility to accommodate the nonstationarity present in power matter spectra 
(which vary in smoothenss across wavenumbers), provide effective uncertainty
quantification (UQ), and allow for the incorporation of expert knowledge 
regarding the wavenumbers over which various outputs are trusted.  To this end, 
we propose a Bayesian hierarchical model which treats functional simulation 
outputs as realizations of a deep Gaussian process \citep[DGP;][]{damianou2013deep} 
centered on the true underlying spectra.  The hierarchical structure naturally
accommodates the correlated observations.  The depth of the GP offers nonstationary
flexibility.  The Bayesian framework facilitates UQ and the incorporation of prior
knowledge.  While Bayesian DGPs have been previously deployed for computer experiments 
with scalar responses \citep[e.g.,][]{sauer2023active,sauer2023vecchia,ming2023deep}, 
they have yet to be extended to correlated functional outputs. 
We will demonstrate profiency on synthetic examples and a low fidelity
cosmological simulation, before deploying our model on the motivating 
Mira-Titan simulation.

@@

For a comprehensive assessment of our Bayesian hierarchical DGP, we benchmark
its performance in two additional settings: with synthetic data mimicing 
the Mira-Titan data and with real-world data from 
the Code for Anisotropies in the Microwave Background \citep[CAMB,][]{lewis2011CAMB}.
The CAMB model also exhibits similar behavior to Mira-Titan, but crucially
provides a ``true" power spectrum corresponding to each cosmology.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth, trim=0 13 0 45, clip=TRUE]{plot_data.png}
    \caption{{\it Left:} The perturbation theory (solid green), low resolution runs (solid grey), 
    and high resolution run  (solid pink) for a particular cosmology. 
    The weighted average of these is shown in dashed orange. Dotted lines at the bottom 
    indicate where each data type is deemed approximately unbiased. 
    {\it Middle:} Same as left, but restricted to wavenumber ($k$) values where the low resolution 
    runs are approximately unbiased.
    {\it Right:} Same as left, but with estimated posterior mean subtracted.
    \textcolor{red}{Rerun on Windows for $\mathcal{P}$}}
    \label{fig:plot_data}
\end{figure}

These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid, we see each low-resolution and the high-resolution run oscillate around 
a mean term, which is relatively smooth, compared to these realizations. When such a 
mean term is subtracted, we will obtain a plot like the right panel in 
Figure \ref{fig:plot_data}.

From the multiple low- and high-resolution spectra across all cosmologies, 
\cite{moran2023mira} obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output. 
We can easily incorporate these estimates in our model to account for nonstationarity 
(i.e., non-constant variance) of the runs. We see as $k$ increases, the runs' precision 
increases in the middle panel of Figure \ref{fig:plot_data}). 
Given that the perturbation theory spectra is non-stochastic, we treat this as having a 
precision of $10^5$. We leverage these precision estimates within our model at each 
data type's unbiased wavenumber values to define the following diagonal precision matrix:
% $\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
% $\Lambda_{\ell} = \text{diag}\left(16\times(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
% $\Lambda_h = \text{diag}\left(c\times(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$.

\[
\Lambda = 
\begin{bmatrix}
\Lambda_p & 0 & 0 \\
0 & \Lambda_\ell & 0 \\
0 & 0 & \Lambda_h
\end{bmatrix}
=
\begin{bmatrix}
\operatorname{diag}(10^5, \ldots, 10^5) &  &  \\
 & \operatorname{diag}(16\times(p_1, \ldots, p_n)) &  \\
 &  & \operatorname{diag}(c\times(p_1, \ldots, p_n))
\end{bmatrix}
\]

\[
\text{where } 
\begin{cases}
\Lambda_p \text{ for } k < 0.04 \\
\Lambda_\ell \text{ for } 0.04 \leq k < 0.25 \\
\Lambda_h \text{ for } 0.25 \leq k \leq 5
\end{cases}
\]

Above, each matrix element is either the corresponding precision value or zero, 
depending on whether the $k$ value is approximately unbiased for that data type. 
With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology. On its own, the weighted average
has too much variation baked in from the limited number of model runs. There is also
no spatial dependence accounting for the correlated errors seen in Figure \ref{fig:plot_data}.
To account for these modeling challenges, we need a more complex model.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra (and quantify uncertainty) for a particular cosmology.  
We focus solely on one cosmology, independent of the others.

@@

Here, we introduce the Bayesian Gaussian Process (GP) model leveraging the aggregated 
data $D_n \equiv \{X_n, \bar Y, \Lambda\}$ for a particular cosmology. We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$.

>>>>

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{amsmath, amsfonts, amssymb, mathrsfs, dsfont}
\usepackage{dcolumn}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{filemod}
\usepackage{natbib}
\usepackage[ruled, vlined]{algorithm2e}
\usepackage{floatrow}
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{bbm} % for indicator function
\usepackage{xcolor} % only for coloring text while revising

\usepackage{tikz} % for pretty diagonal matrix
\usetikzlibrary{matrix,decorations.pathreplacing,calc}

@@

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the power matter spectra 
is rather complex.  Although we presume there exists some true matter spectra as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation outputs 
a convoluted patchwork of potential spectra, which vary in fidelity and are trusted
only over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all available simulation information into an effective estimate of the true underlying
matter spectrum.

There are several challenges to estimating the underlying spectra.  Our model
must: handle multiple correlated functional observations, offer enough 
flexibility to accommodate the nonstationarity present in power matter spectra 
(which vary in smoothenss across wavenumbers), provide effective uncertainty
quantification (UQ), and allow for the incorporation of expert knowledge 
regarding the wavenumbers over which various outputs are trusted.  To this end, 
we propose a Bayesian hierarchical model which treats functional simulation 
outputs as realizations of a deep Gaussian process \citep[DGP;][]{damianou2013deep} 
centered on the true underlying spectra.  The hierarchical structure naturally
accommodates the correlated observations.  The depth of the GP offers nonstationary
flexibility.  The Bayesian framework facilitates UQ and the incorporation of prior
knowledge.  While Bayesian DGPs have been previously deployed for computer experiments 
with scalar responses \citep[e.g.,][]{sauer2023active,sauer2023vecchia,ming2023deep}, 
they have yet to be extended to correlated functional outputs. 
To demonstrate the profiency of our Bayesian hierarchical DGP, 
we benchmark its performance in two unique settings: with synthetic data 
mimicing the Mira-Titan data and with real-world data from 
the Code for Anisotropies in the Microwave Background \citep[CAMB;][]{lewis2011CAMB}.
The CAMB model also exhibits similar behavior to Mira-Titan, but crucially
provides a ``true" power spectrum corresponding to each cosmology.

@@

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth, trim=0 13 0 45, clip=TRUE]{plot_data.png}
    \caption{{\it Left:} The perturbation theory (solid green), low resolution runs (solid grey), 
    and high resolution run  (solid pink) for a particular cosmology. 
    The weighted average of these is shown in dashed orange. Dotted lines at the bottom 
    indicate where each data type is deemed approximately unbiased. 
    {\it Middle:} Same as left, but restricted to wavenumber ($k$) values where the low resolution 
    runs are approximately unbiased.
    {\it Right:} Same as left, but with estimated posterior mean subtracted.
    \textcolor{red}{Rerun on Windows for $\mathcal{P}$.  Make axis/labels and legend 
    a bit larger so they are easier to read.  Am I tripping or are the horizontal lines
    in the left panel not lined up correctly?}}
    \label{fig:plot_data}
\end{figure}

These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), we see each low-resolution and the high-resolution run oscillate around 
a mean term, which is relatively smooth, compared to these realizations. When this
mean term is subtracted, we will obtain a plot like the right panel in 
Figure \ref{fig:plot_data} \textcolor{red}{(This plot goes back to the original range,
would it be prettier and easier for the reader to follow if we kept it to the 
range of the middle panel here?)}

Notice, in the center panel of Figure \ref{fig:plot_data}, the precision of each
spectra increases as $k$ increases.  The curves get progressively tighter from left
to right.  We would like to conveniently wrap this domain-specific knowledge inside
our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output.
To streamline notation, we denote three unique diagonal matrices:
\[
\Lambda_p = 
   \begin{tikzpicture}[decoration={brace,amplitude=5pt},baseline=(current bounding box.west)]
     \matrix (magic) [matrix of math nodes,left delimiter=(,right delimiter=)] {
      10^5 \\
      & \ddots \\
      & & 10^5 \\
      & & & 0 \\
      & & & & 0 \\
      & & & & & \ddots \\
      & & & & & & 0\\
     };
     \draw[decorate] (magic-1-1.north) -- (magic-3-3.north east) node[above=5pt,midway,sloped] {$k < 0.04$};
   \end{tikzpicture}
\quad
\Lambda_l = 
   \begin{tikzpicture}[decoration={brace,amplitude=5pt},baseline=(current bounding box.west)]
     \matrix (magic) [matrix of math nodes,left delimiter=(,right delimiter=)] {
      0 \\
      & \ddots \\
      & & 0 \\
      & & & 10p_1 \\
      & & & & \ddots \\
      & & & & & 10p_n ? \\
      & & & & & & 0 \\
      & & & & & & & \ddots \\
      & & & & & & & & 0 \\
     };
     \draw[decorate] (magic-4-4.north) -- (magic-6-6.north east) node[above=5pt,midway,sloped] {$0.04\leq k < 0.25$};
   \end{tikzpicture}
\]
\[
\Lambda_h = 
   \begin{tikzpicture}[decoration={brace,amplitude=5pt},baseline=(current bounding box.west)]
     \matrix (magic) [matrix of math nodes,left delimiter=(,right delimiter=)] {
      0 \\
      & \ddots \\
      & & 0 \\
      & & & cp_1 \\
      & & & & \ddots \\
      & & & & & cp_n \\
     };
     \draw[decorate] (magic-4-4.north) -- (magic-6-6.north east) node[above=5pt,midway,sloped] {$0.25\leq k \leq 5$};
   \end{tikzpicture}
\]
\textcolor{red}{(@Steve - I am still quite confused here.  Are the above matrices helpful?
I'm confused by $p_i$.  Isn't there some overlap
between the low and high res runs?  The piecewise form you had doesn't have any overlap.  You
had the large $\Lambda$ as the concatenation of the matrices, but then you also had
it as a sum.  Perhaps we should talk this out sometime.  There has to be an easier 
way to convey how we calculate the weighted average.  Perhaps we could represent 
this better with indicator functions?  
Or as a function of $k_i$?  The equations below are just some ideas.)}
\[
\Lambda_p(k) = 10^5 \mathbbm{I}_{\{k < 0.04\}}
\]
where $\mathbbm{I}$ is the indicator function.
\[
\Lambda(k_i) = \begin{cases}
    10^5 \text{ for } k_i < 0.04 \text{  (perturbation theory)}\\
    16p_i \text{ for } 0.04 \leq k_i < 0.25 \text{  (low resolution)}\\
    cp_i \text{ for } 0.25 \leq k_i \leq 5 \text{  (high resolution)}
    \end{cases},
\]
where we use the sufficiently high precision of $10^5$ to impose the deterministic
nature of the perturbation theory output.  \textcolor{red}{(Then comment on how 
we can put this all into one big diagonal matrix that we multiply to our y values?)}

% $\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
% $\Lambda_{\ell} = \text{diag}\left(16\times(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
% $\Lambda_h = \text{diag}\left(c\times(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$.

%\[
%\Lambda = 
%\begin{bmatrix}
%\Lambda_p & 0 & 0 \\
%0 & \Lambda_\ell & 0 \\
%0 & 0 & \Lambda_h
%\end{bmatrix}
%=
%\begin{bmatrix}
%\operatorname{diag}(10^5, \ldots, 10^5) &  &  \\
% & \operatorname{diag}(16\times(p_1, \ldots, p_n)) &  \\
% &  & \operatorname{diag}(c\times(p_1, \ldots, p_n))
%\end{bmatrix}
%\]

\textcolor{red}{(This part will need some revising, it is still unclear to me
how the weighted average is calculated.  Although I know in reality it is super simple.}
Above, each matrix element is either the corresponding precision value or zero, 
depending on whether the $k$ value is approximately unbiased for that data type. 
With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology. On its own, the weighted average
has too much variation baked in from the limited number of model runs. There is also
no spatial dependence accounting for the correlated errors seen in Figure \ref{fig:plot_data}.
To account for these modeling challenges, we need a more complex model.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology,
independent of the others.  Let $D_n = \{X_n, \bar{Y}, \Lambda\}$ represent the aggregated 
data for a particular cosmology.  Let $S$ represent the true (but unknown) underlying
power matter spectra from which $D_n$ was generated.
\textcolor{red}{(What is X?  Why is $\bar{Y}$ capital here, but it wasn't earlier?
Since $\Lambda$ is fixed and doesn't change, I do not think we need to consider it part
of the data.)}

@@

Here, we introduce the Bayesian Gaussian Process (GP) model We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$.
